Box 1 - TDT Curves

### 2023-09-27 ms/ms.qmd
Remove within species We really don't have a lot

---
title: ""
bibliography: ../bib/refs.bib
csl: ../bib/nature.csl
format:
  docx:
    reference-doc: ../bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

Parameters from Thermal Death Time (TDT) curves provide a powerful tool for understanding the thermal sensitivity of organisms. But there are drawbacks to estimating TDT curves in comparison to other lethal (e.g. $CT_{max}$) and sublethal (CFL) measures. TDT curves are logistically challenging requiring more individuals for reliable estimates than other measures like $CT_{max}$ and CFL’s. TDT curves (and $CT_{max}$) also cannot be used for all organisms due to ethical and conservation considerations. Ethical issues could be alleviated by the use of sublethal measures of thermal tolerance like CFL’s (ref), but we do not know how well lethal (TDT curves/$CT_{max}$) and sublethal (CFL’s) measures capture thermal sensitivities/vulnerabilities. Identifying whether a correlation exists between lethal and sub-lethal limits will help us understand the mechanisms of vulnerability to heat stress that might lead to population extinction well before mortality is identified. Because of these limitations data on thermal sensitivity/heat vulnerability is missing for many species, creating significant knowledge gaps across the tree of life.

Phylogenetic comparative approaches could be leveraged to help fill these knowledge gaps because closely related species are often more likely to share similar traits owing to their evolutionary history (phylogenetic niche conservatism) (ref). If heat tolerance is linked to the phylogeny, shared characteristics that make species more or less heat tolerant will not be randomly distributed across the phylogeny but biased by species groups (ref). We can can use evolutionary relatedness to predict species thermal sensitivities and such insights can be used to inform policy and conservation related decisions making for species with little to no data.

@@

We integrated static lethal limits ($CT_{max}$) collected by Heerwaarden & Sgrò @Heerwaarden2021 and Kellerman *et al* @Kellerman2012 with dynamic lethal limit measures from TDT curces [i.e., slope ($z^{TDT}$) and intercept ($CT_{max}^{TDT}$)] estimated by Jorgensen *et al* @Jorgensen2019 and . We also collected data to estimate TDT curve parameters for eight additonal species following protocols outlined in @Orsted2022. Critical fertility limits for *Drosophilids* were compiled from @Parratt2021 and @Heerwaarden2021. These data were integrated with data on lethal limits at the species-level. Overall, we obtained data for `r num_spp[1]` species ($N_{CT_{max}}$ = `r num_spp[2]` species; $N_{z^{TDT}}$ = `r num_spp[4]` species and $N_{CFLs}$ = `r num_spp[3]` species). When the same species were sampled multiple times we retained within species variation in models. We then fit the following model in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`)[@R2023] with the package *MCMCglmm* (vers. `r utils::packageVersion("MCMCglmm")`), which uses data augmentation during model fittng to estimate missing data [@Hadfield2010]:

@@

 Models including phylogeny were better supported than models without phylogeny ($\Delta DIC$ = `r mcmcglmm_model_z_trans_nophylo$DIC - mcmcglmm_model_z_trans$DIC`), with all traits exhibiting a moderate to strong phylogenetic signal indicating evolutionary conservatism in heat tolerance (Phylogenetic $H^2$: $CT_{max}$ – `r posterior.mode(h2_ctmax)`, 95% CI: `r HPDinterval(h2_ctmax)[,"lower"]` to `r HPDinterval(h2_ctmax)[,"upper"]`; $z^{TDT}$ – `r posterior.mode(h2_z)`, 95% CI: `r HPDinterval(h2_z)[,"lower"]` to `r HPDinterval(h2_z)[,"upper"]`; CFLs – `r posterior.mode(h2_cfls)`, 95% CI: `r HPDinterval(h2_cfls)[,"lower"]` to `r HPDinterval(h2_cfls)[,"upper"]`). While the model also suggested that there was within species variation, missing data and lack of replicate species populations make conclusions premature ($I^2_{species}$: $CT_{max}$ – `r posterior.mode(spp_ctmax)`, 95% CI: `r HPDinterval(spp_ctmax)[,"lower"]` to `r HPDinterval(spp_ctmax)[,"upper"]`; $z^{TDT}$ – `r posterior.mode(spp_z)`, 95% CI: `r HPDinterval(spp_z)[,"lower"]` to `r HPDinterval(spp_z)[,"upper"]`; CFLs – `r posterior.mode(spp_cfls)`, 95% CI: `r HPDinterval(spp_cfls)[,"lower"]` to `r HPDinterval(spp_cfls)[,"upper"]`).

@@

We analysed the thermal death time curves for 8 new species of *Drosophila* using the same methods as @Jorgensen2019, using linear models with log 10 coma time as the response and assay temperature as the predcitor [@fig-tdt]. We then used the model to estimate the slope and intercept ($CT_{max}$) for each species [@tbl-1].

>>>>

---
title: ""
bibliography: ../bib/refs.bib
csl: ../bib/nature.csl
format:
  docx:
    reference-doc: ../bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

Parameters from Thermal Death Time (TDT) curves provide a powerful tool for understanding the thermal sensitivity of organisms. But there are drawbacks to estimating TDT curves in comparison to other lethal (e.g. $CT_{max}$) and sublethal (CFL) measures. TDT curves are logistically challenging requiring more individuals for reliable estimates than other measures like $CT_{max}$ and CFL’s. TDT curves (and $CT_{max}$) also cannot be used for all organisms due to ethical and conservation considerations. Ethical issues could be alleviated by the use of sublethal measures of thermal tolerance like CFL’s [@Wang2022], but we do not know how well lethal (TDT curves/$CT_{max}$) and sublethal (CFL’s) measures capture thermal sensitivities/vulnerabilities. Identifying whether a correlation exists between lethal and sub-lethal limits will help us understand the mechanisms of vulnerability to heat stress that might lead to population extinction well before mortality is identified. Because of these limitations data on thermal sensitivity/heat vulnerability is missing for many species, creating significant knowledge gaps across the tree of life.

Phylogenetic comparative approaches could be leveraged to help fill these knowledge gaps because closely related species are often more likely to share similar traits owing to their evolutionary history (phylogenetic niche conservatism) [@Losos2008]. If heat tolerance is linked to the phylogeny, shared characteristics that make species more or less heat tolerant will not be randomly distributed across the phylogeny but biased by species groups [@McKinney1997]. We can use evolutionary relatedness to predict species thermal sensitivities and such insights can be used to inform policy and conservation related decisions making for species with little to no data.

@@

We integrated static lethal limits ($CT_{max}$) collected by Heerwaarden & Sgrò @Heerwaarden2021 and Kellerman *et al* @Kellerman2012 with dynamic lethal limit measures from TDT curces [i.e., slope ($z^{TDT}$) and intercept ($CT_{max}^{TDT}$)] estimated by Jorgensen *et al* @Jorgensen2019 and Rezende *et al* @Rezende2014. We also collected data to estimate TDT curve parameters for eight additonal species following protocols outlined in Jorgensen *et al* @Orsted2022 and Rezende *et al* @Rezende2014 (See *Supplementary Materials*). Critical fertility limits for *Drosophilids* were compiled from @Parratt2021 and @Heerwaarden2021. These data were integrated with data on lethal limits at the species-level. Overall, we obtained data for `r num_spp[1]` species ($N_{CT_{max}}$ = `r num_spp[2]` species; $N_{z^{TDT}}$ = `r num_spp[4]` species and $N_{CFLs}$ = `r num_spp[3]` species). When the same species were sampled multiple times we retained within species variation in models. We then fit the following model in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`)[@R2023] with the package *MCMCglmm* (vers. `r utils::packageVersion("MCMCglmm")`), which uses data augmentation during model fittng to estimate missing data [@Hadfield2010]:

@@

 Models including phylogeny were better supported than models without phylogeny ($\Delta DIC$ = `r mcmcglmm_model_z_trans_nophylo$DIC - mcmcglmm_model_z_trans$DIC`), with all traits exhibiting a moderate to strong phylogenetic signal indicating evolutionary conservatism in heat tolerance (Phylogenetic $H^2$: $CT_{max}$ – `r posterior.mode(h2_ctmax)`, 95% CI: `r HPDinterval(h2_ctmax)[,"lower"]` to `r HPDinterval(h2_ctmax)[,"upper"]`; $z^{TDT}$ – `r posterior.mode(h2_z)`, 95% CI: `r HPDinterval(h2_z)[,"lower"]` to `r HPDinterval(h2_z)[,"upper"]`; CFLs – `r posterior.mode(h2_cfls)`, 95% CI: `r HPDinterval(h2_cfls)[,"lower"]` to `r HPDinterval(h2_cfls)[,"upper"]`).

@@

We analysed the thermal death time curves for 8 new species of *Drosophila* using the same methods as @Jorgensen2019, using linear models with log 10 coma time as the response and assay temperature as the predcitor (@fig-tdt). We then used the model to estimate the slope and intercept ($CT_{max}$) for each species (Table 1).
